Tree height, microhabitat, and hydraulic traits shape drought responses in a temperate broadleaf forest

### 2019-10-12 manuscript/climate_sensitivity_draft.Rmd
Tried fixing errors with latex

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("tables_figures/hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(hypothesis_table, booktabs = TRUE, caption = "**Table 1. Summary of hypotheses, corresponding specific predictions, and results.** We count predictions as fully supported / rejected when the response matches/contradicts the prediction in both univariate and multivariate models (when applicable). Parentheses indicate that predictions were partially supported/ rejected--i.e., that the direction of response matched/contradicted the prediction in some but not all models.") %>%
  pack_rows("H1.0. Larger-diameter trees have lower drought resistance.", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H1.1. Tall trees have lower drought resistance.", 2, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H1.2. Trees with more exposed crowns have lower drought resistance .", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H1.3. Small trees (lower root volume) suffer more in drier microhabitats.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H2.1. Species traits predict drought resistance.", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
  pack_rows("H2.2. Taller trees have more drought-resistant traits.", 11, 15, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H2.3. Size-dependent drought resistance is not driven by functional traits.", 16, 16, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H3.1. Community drought resistance differs across droughts.", 17, 17, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
  pack_rows("H3.2. The direction of responses to predictor variables differs across droughts.", 18, 18, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H3.3. The strength of responses to predictor variables vary across droughts.", 19, 19, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
  column_spec(1, width = "12cm") %>%
  column_spec(2:5, width = "1cm") %>%
  column_spec(6, width = "2cm") %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5) %>%
  kableExtra::landscape()
```

@@

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("tables_figures/variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)

kable(hypothesis_table, booktabs = TRUE, caption = "**Table 2. Summary of variables**", format = 'latex', escape = TRUE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  add_header_above(c(" "," "," "," "," "," ", "observed values" = 3, " ")) %>%
  column_spec(4, width = "7cm") %>%
  pack_rows("Dependent variable", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("Independent variables", 2, 18, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("tree size", 5, 6, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("microhabitat", 7, 11, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("species' traits", 12, 18, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  kable_styling()
```

@@

```{r, Table3, eval=TRUE, echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
table2 <- read.csv("tables_figures/species_table.csv")

@@

kable(table2, booktabs=TRUE, caption = "**Table 3. Overview of analyzed species, their productivity in the plot, numbers and sizes sampled, and traits.**  Given are DBH mean and range of cored trees, the number of cores represented by each crown position of each species, and mean hydraulic trait measurements. Units of measurements are in mm (DBH), \\% (PLA), g/m2 (LMA), MPa (TLP), and g/cm3 (WD).") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

@@

Linear mixed models were run following the order of the hypotheses as seen in Figure ??? [individual_tested_traits]. Using the [@R-pointRes] package, we set up models with the resistance value as the response variable (excluding outliers with values of $R\ge 2$), and each prediction's variable as the independent variable. Variables' importance in predicting drought tolerance was calculated from mixed-effects models and the lowest AICc [@R-lme4, @R-AICcmodavg].Null models were determined in order of the predictions. First, we analyzed the combined scenario to determine if "year" was significant. Upon establishing this, we tested height and DBH as size parameters. Although both were significant, height was kept due to its larger delta AICc compared with the null model. We then tested the remaining biophysical and hydraulic traits individually against a null model containing height and year. This yielded Figure ??? (cand_full). All variables with dAICc >2 were used as candidates for each scenario's best model (figure ???? (tested_traits_best))

@@

In the 60-year period between 1950 and 2009, there were three droughts that met our criteria of anomalously dry climatic conditions coupled with substantial reductions in tree growth for at least some portions of the community: 1966, 1977, and 1999 (Fig. 1). 
We excluded one year (1991) meeting the growth reduction criteria (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not among the strongest droughts of the study period (**DETAILS**). Rather, the severity of growth reduction may be explained by defoliation by gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which most stronly impacted *Quercus* spp. (*Cite Shenanadoah paper, if accepted*). 
Climatically, these droughts included three of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August, which are the months to which annual tree growth at this site is most sensitive [@helcoski_growing_2019]. Specifically, 1966, 1977, and 1999 had mean MJJA PET-PRE of 83.37, 86.97, and 80 mm mo<sup>-1</sup>, respectively. The years 1964 and 2007 also ranked in the top five driest (PET-PRE =83.87 and 82.13 mm mo<sup>-1</sup>), but *were not among the lowest in terms of PDSI and were not identified as a pointer yeasr.* 
**The droughts differed in timing/duration/etc. .. The year 1966 was preceded by two relatively dry years... 1964 among five driest in terms of May-August [PET-PRE],  1965  also anomalously hot and dry. **

Community-level tree growth responses to these droughts were modest, with modal resistance values of #, #, and # for 1966, 1977, and 1999, respectively (Fig. 1b). In each drought, roughly 30% of the cored trees suffered $\ge$ 30% growth reductions ($R$ $\le$ 0.7): #% in 1966, #2% un 1977, and #% in 1999. *Some* trees exhibited increased growth: ($R$ > 1.0): #% in 1966, #% un 1977, and #% in 1999.

@@

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("tables_figures/multivariate_models_table.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
kable(hypothesis_table, booktabs = TRUE, caption = "**Table 5. Summary of R^2 and coefficients of the best multivariate models for each drought instance.** Models are ranked by AIC, and we show all models whose AIC value falls within 2.0 of the best model (dAICc<2).", format = 'latex', escape = TRUE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  add_header_above(c(" "," "," "," "," ", "crown position" = 4, " ","xylem architecture" = 3," ", " ")) %>%
  pack_rows("", 1, 6, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 7, 11, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
  pack_rows("", 12, 15, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
  pack_rows("", 16, 24, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
  kable_styling()
```

>>>>

```{r Table 1, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("tables_figures/hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(hypothesis_table, booktabs = TRUE, caption = "**Table 1. Summary of hypotheses, corresponding specific predictions, and results.** We count predictions as fully supported / rejected when the response matches/contradicts the prediction in both univariate and multivariate models (when applicable). Parentheses indicate that predictions were partially supported/ rejected--i.e., that the direction of response matched/contradicted the prediction in some but not all models.") %>%
  pack_rows("H1.0. Larger-diameter trees have lower drought resistance.", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H1.1. Tall trees have lower drought resistance.", 2, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H1.2. Trees with more exposed crowns have lower drought resistance .", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H1.3. Small trees (lower root volume) suffer more in drier microhabitats.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H2.1. Species traits predict drought resistance.", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
  pack_rows("H2.2. Taller trees have more drought-resistant traits.", 11, 15, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H2.3. Size-dependent drought resistance is not driven by functional traits.", 16, 16, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H3.1. Community drought resistance differs across droughts.", 17, 17, latex_gap_space = "1em", colnum = 1, hline_before = TRUE) %>%
  pack_rows("H3.2. The direction of responses to predictor variables differs across droughts.", 18, 18, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("H3.3. The strength of responses to predictor variables vary across droughts.", 19, 19, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
  column_spec(1, width = "12cm") %>%
  column_spec(2:5, width = "1cm") %>%
  column_spec(6, width = "2cm") %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5) %>%
  kableExtra::landscape()
```

@@

```{r Table 2, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
table2 <- read.csv("tables_figures/variables_table.csv", stringsAsFactors = FALSE, check.names = FALSE)

kable(table2, booktabs = TRUE, caption = "**Table 2. Summary of variables**", format = 'latex', escape = FALSE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  add_header_above(c(" "," "," "," "," "," ", "observed values" = 3, " ")) %>%
  column_spec(4, width = "7cm") %>%
  pack_rows("Dependent variable", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("Independent variables", 2, 18, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("tree size", 5, 6, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("microhabitat", 7, 11, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("species' traits", 12, 18, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  kable_styling()
```

@@

```{r, Table3, eval=TRUE, echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
table3 <- read.csv("tables_figures/species_table.csv")

@@

kable(table3, booktabs=TRUE, caption = "**Table 3. Overview of analyzed species, their productivity in the plot, numbers and sizes sampled, and traits.**  Given are DBH mean and range of cored trees, the number of cores represented by each crown position of each species, and mean hydraulic trait measurements. Units of measurements are in mm (DBH), \\% (PLA), g/m2 (LMA), MPa (TLP), and g/cm3 (WD).") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

@@

Linear mixed models were run following the order of the hypotheses as seen in Figure ??? [individual_tested_traits]. Using the [@R-pointRes] package, we set up models with the resistance value as the response variable (excluding outliers with values of $R\ge 2$), and each prediction's variable as the independent variable. Variables' importance in predicting drought tolerance was calculated from mixed-effects models and the lowest AICc [@R-lme4, @R-AICcmodavg].Null models were determined in order of the predictions. First, we analyzed the combined scenario to determine if "year" was significant. Upon establishing this, we tested height and DBH as size parameters. Although both were significant, height was kept due to its larger delta AICc compared with the null model. We then tested the remaining biophysical and hydraulic traits individually against a null model containing height and year. This yielded Figure ??? (candfull). All variables with dAICc > 2 were used as candidates for each scenario's best model (figure ???? (tested traits best))

@@

In the 60-year period between 1950 and 2009, there were three droughts that met our criteria of anomalously dry climatic conditions coupled with substantial reductions in tree growth for at least some portions of the community: 1966, 1977, and 1999 (Fig. 1). 
We excluded one year (1991) meeting the growth reduction criteria (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) because this year was not among the strongest droughts of the study period (**DETAILS**). Rather, the severity of growth reduction may be explained by defoliation by gypsy moths (*Lymantria dispar* L.) from approximately 1988-1995, which most stronly impacted *Quercus* spp. (*Cite Shenandoah paper, if accepted*). 
Climatically, these droughts included three of the five years between 1950 and 2009 with greatest moisture deficit (PET-PRE) during the peak growing season months of May-August, which are the months to which annual tree growth at this site is most sensitive [@helcoski_growing_2019]. Specifically, 1966, 1977, and 1999 had mean MJJA PET-PRE of 83.37, 86.97, and 80 mm mo<sup>-1</sup>, respectively. The years 1964 and 2007 also ranked in the top five driest (PET-PRE =83.87 and 82.13 mm mo<sup>-1</sup>), but *were not among the lowest in terms of PDSI and were not identified as a pointer yeasr.* 
**The droughts differed in timing/duration/etc. .. The year 1966 was preceded by two relatively dry years... 1964 among five driest in terms of May-August [PET-PRE],  1965  also anomalously hot and dry. **

Community-level tree growth responses to these droughts were modest, with modal resistance values of #, #, and # for 1966, 1977, and 1999, respectively (Fig. 1b). In each drought, roughly 30% of the cored trees suffered $\ge$ 30% growth reductions ($R$ $\le$ 0.7): #% in 1966, #2% un 1977, and #% in 1999. *Some* trees exhibited increased growth: ($R$ > 1.0): #% in 1966, #% in 1977, and #% in 1999.

@@

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("tables_figures/multivariate_models_table.csv", stringsAsFactors = FALSE, check.names = FALSE) #note: this table is a rearranged version of `tested_traits_best_coeff.csv`.
kable(hypothesis_table, booktabs = TRUE, caption = "**Table 5. Summary of R\\^2 and coefficients of the best multivariate models for each drought instance.** Models are ranked by AIC, and we show all models whose AIC value falls within 2.0 of the best model (dAICc<2).", format = 'latex', escape = TRUE) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12, protect_latex = TRUE) %>%
  add_header_above(c(" "," "," "," "," ", "crown position" = 4, " ","xylem architecture" = 3," ", " ")) %>%
  pack_rows("", 1, 6, latex_gap_space = ".4em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("", 7, 11, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
  pack_rows("", 12, 15, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
  pack_rows("", 16, 24, latex_gap_space = ".4em", colnum = 1, hline_before = TRUE) %>%
  kable_styling()
```
